---
title: "ggplot"
author: "Prof. Harbert"
date: "September 19, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ggplot

The ggplot library is part of the 'tidyverse'. While not the only way to make great graphics in R, ggplot is the leading data visualization software in R.

More: https://ggplot2.tidyverse.org/ 

And a great cheat-sheet: https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf 

```{r}
# install.packages("tidyverse") # if not yet installed

library(tidyverse)
```

```{r}


demo <- read.delim("https://raw.githubusercontent.com/rsh249/bioinformatics/master/data/demo.tab") 

# examine
head(demo) # view the first 6 rows
head(demo, n = 10) # view the first n rows
str(demo) # describe the STRucture of the data.frame


# generate summary statistics
summary(demo$x)

# SCATTERPLOTS
ggplot(data = demo) +
  geom_point(mapping = aes(x = x, y = y))

# change x and y axis titles
ggplot(data = demo) +
  geom_point(mapping = aes(x = x, y = y)) +
  xlab("time") + ylab("Temperature (C)")

# use transparency to better visualize clusters of points
# with the "alpha" parameter

ggplot(data = demo) +
  geom_point(mapping = aes(x = x, y = y,alpha =0.5)) +
  xlab("time") + ylab("Temperature (C)")


#or color by a group column
ggplot(data = demo) +
  geom_point(mapping = aes(x = x, y = y, col=gp)) +
  xlab("time") + ylab("Temperature (C)")
```

# Aesthetics:

Depending on the graphics you are building you can define the aes() function in the ggplot() function OR (as above) in the the geom function call:

```{r}
ggplot(data = demo, aes(x=x, y=y, col=gp)) +
  geom_point() + xlab("time") + ylab("Temperature (C)")
```

# Saving plot objects:
You can save a basic plot object by assigning it to an R object:

```{r}
base = ggplot(data = demo, aes(x=x, y=y, col=gp)) +
  geom_point() + xlab("time") + ylab("Temperature (C)")
```

This object ('base' in this example) can then be reused and modified. We will cover several modifications below.

# Facetting -- multiple plots

In ggplot the facet (facet_wrap, facet_grid) functions will make a multipanel plot with subplots sorted by a grouping factor.

```{r}
base +
  facet_wrap(~ gp)

```

# Style, Colors, and Themes

The basic ggplot formatting is OK, but we can do better with a few stylistic tweaks.

## Themes

One of the easiest is to attach a theme function to the end of your ggplot call. These can be viewed with: ?theme_bw()


e.g., theme_bw()
```{r}
base +
  facet_wrap(~ gp) +
  theme_bw()

```

e.g., theme_minimal()
```{r}
base +
  facet_wrap(~ gp) +
  theme_minimal()

```

I like this version so I'm going to save it as another R object that keeps this formatting. 

```{r}
minimal = base + 
  facet_wrap(~ gp) +
  theme_minimal()
```

## Color palettes

With R Color Brewer 

```{r}

library(RColorBrewer)
display.brewer.all()

```


And if we want to call one of these in ggplot. Use + scale_color_brewer('palette name')

e.g., with 'Dark2' 
```{r}

minimal + 
  scale_color_brewer(palette="Dark2")

```

## Manual colors and outside palettes:

The ggplot scale_color_manual() lets you pass any other vector of colors to your ggplot call. This lets you define colors: (WARNING: these can be really bad!) 

```{r}
minimal + scale_color_manual(values=c('blue', 'orange', 'green'))
```



Or grab colors from outside libraries: e.g., the wesanderson library:

```{r}
library(wesanderson) #colors from Wes Anderson Movies
minimal + scale_color_manual(values=wes_palette(n=3, name = "GrandBudapest1"))
```

More on colors and coloring functions can be found here:

http://www.sthda.com/english/wiki/ggplot2-colors-how-to-change-colors-automatically-and-manually


# Other plot types:

## Boxplots 
```{r}
ggplot(data = demo, aes(x = sub, y = y)) +
  geom_boxplot()
```

## Barplots 
```{r}
ggplot(data = demo, aes(sub)) +
  geom_bar()
```

## Histograms
```{r}
ggplot(data=demo, aes(y)) + geom_histogram(binwidth = 5, col ='white')

ggplot(data=demo, aes(y)) + geom_histogram(binwidth = 5, col ='white') + facet_wrap(~gp)
```


And many, many, more:

https://r4stats.com/examples/graphics-ggplot2/


# Homework:

Using ggplot (and all of the tips and resources here), make one or more graphic representations of your group's data and post it to [#ggplot](https://bio200-bioinformatics.slack.com/messages/CCY83KSS2).




