---
title: "GBIF, iNaturalist, and Primary Biodiversity Data"
author: "Prof. Harbert"
date: "November 14, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Primary Biodiversity Data

Observations of the occurrence of a species are the fundamental unit of biodiversity data. We will explore in this unit where to look for open-access occurrence data, how to access those sources from R, and tools for visualizing point distributions of species.

### Library 'spocc'

A great tool from the rOpenSci consortium (a group of developers building R capacity for open science).

Package details on [GitHub](https://github.com/ropensci/spocc)

Tutorial [here](https://ropensci.org/tutorials/spocc_tutorial/)

We should all have spocc installed, but if not try:

```{r, eval=FALSE}

install.packages('spocc')

```


With spocc installed we can try a simple query of the GBIF database that we have seen briefly before. 

```{r}
library(spocc)
ulmus <- occ(query='Ulmus americana', from='gbif')
```

The data are returned as an "S3 class" object. Somewhere in there is a tidyverse tibble (like a table but not).

```{r, eval=FALSE}
print(ulmus) ## Not obvious what or where the data are
```

```{r, eval=FALSE}
View(ulmus)
```

Maybe it's still not obvious how we get in. To view an element of the data returned we use the "$" operator and call each by name. In general it's easier to convert these to regular R data frame objects since not everything we want to do with these data is compatible with the tidyverse/spocc formatting.

```{r}

df = as.data.frame(occ2df(ulmus$gbif))

#Also try:
#head(df)
#colnames(df) #!! That's a lot of columns!!

```


### mapr: Leaflet mapping of species distribution data.

To create interactive graphics showing species occurrence locations and some metadata we can use 'mapr'. This library uses a JavaScript library known as leaflet and Open Street Maps services (and others!) to create interactive maps that you can navigate through and click on points to pop-up metadata about each occurrence.

If not already done:
```{r, eval=FALSE}
install.packages('mapr')
```

Then call map_leaflet() either on the spocc object:

```{r}
library(mapr)
map_leaflet(ulmus)

```

OR with the data.frame:

```{r}
map_leaflet(df)
```


'mapr' shows the data for the first few columns in each pop-up tab. We can control what is shown there by only passing some columns to map_leaflet(). 

```{r}
map_leaflet(df[,c('name', 'longitude', 'latitude', 'stateProvince', 'country', 'year', 'occurrenceID')])
```