---
title: "GBIF, iNaturalist, and Primary Biodiversity Data"
author: "Prof. Harbert"
date: "November 14, 2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Primary Biodiversity Data

Observations of the occurrence of a species are a fundamental unit of biodiversity data. We will explore in this unit, where to look for open-access occurrence data, how to access those sources from R, and tools for visualizing point distributions of species.

### Library 'spocc'

A great tool from the [rOpenSci consortium](https://ropensci.org/) (a group of developers building R capacity for open science).

Package details on [GitHub](https://github.com/ropensci/spocc)

Tutorial [here](https://ropensci.org/tutorials/spocc_tutorial/)

We should all have spocc installed, but if not try:

```{r, eval=FALSE}

install.packages('spocc')

```


With spocc installed we can try a simple query of the GBIF database that we have seen briefly before. 

```{r}
library(spocc)
spdist <- occ(query='Crotalus horridus', from='gbif')


```

The data are returned as an "S3 class" object. Somewhere in there is a tidyverse tibble (like a data frame table, but not).

```{r, eval=FALSE}
print(spdist) ## Not obvious what or where the data are
```

```{r, eval=FALSE}
View(spdist)
```

Maybe it's still not obvious how we get in. To view an element of the data returned we use the "$" operator and call each by name. In general it's easier to convert these to regular R data frame objects since not everything we want to do with these data is compatible with the tidyverse/spocc formatting.

```{r}

df = as.data.frame(occ2df(spdist$gbif))

#Also try:
#head(df)
#colnames(df) #!! That's a lot of columns!!

```


### mapr: Exploratory interactive mapping of species distribution data.

To create interactive graphics showing species occurrence locations and some metadata we can use 'mapr'. This library uses a JavaScript library known as leaflet and Open Street Maps services (and others!) to create interactive maps that you can navigate through and click on points to pop-up metadata about each occurrence.

If not already done:
```{r, eval=FALSE}
install.packages('mapr')
```

Then call map_leaflet() either on the spocc object:

```{r}
library(mapr)
map_leaflet(spdist)

```

OR with the data.frame:

```{r}
map_leaflet(df)
```


'mapr' shows the data for the first few columns in each pop-up tab. We can control what is shown there by only passing some columns to map_leaflet(). 

```{r}
map_leaflet(df[,c('name', 'longitude', 'latitude', 'stateProvince', 'country', 'year', 'occurrenceID')])
```

Specifying columns makes it much easier to sift through large amounts of data to check sources and look for patterns of bias.

NOTE: mapr only works with data formatted by spocc and related libraries.

### More with spocc queries.

Do you notice something odd when you run:

```{r}
nrow(df)
```

Check how many records are returned for the same search on the [GBIF website](https://www.gbif.org/species/5361872) 

Our query only returned the first 500 records because that is the default for the occ() function.

We can fix that:

```{r}
spdist2 <- occ(query='Crotalus horridus', limit=2500)
map_leaflet(spdist2)
```

# Meeting up with WorldClim

R gives us the tools to plot spatial point data like our occurrence records over a geospatial raster object like the WorldClim data. R also gives us tools to access the raster data for our occurrence points.

Load WorldClim again: (If you are in the same folder then you will not have to download the data again so this should be quick).

```{r}
library(raster)
wc = getData('worldclim', var='bio', res = 5)

```
Plot raster with points over:

```{r}
library(ggplot2)
ext = extent(-125, -55, 20, 60)
wc = crop(wc, ext)

wc_df = as.data.frame(wc, xy=TRUE)
sp_df = occ2df(spdist2)
ggplot() +
  geom_raster(data = wc_df, aes(x = x, y = y, fill = bio1/10)) +
  geom_point(data=sp_df, aes(x=longitude, y=latitude), col='green') +
  coord_quickmap() +
  theme_bw() + 
  scale_fill_gradientn(colours=c('navy', 'white', 'darkred'),
  na.value = "black")


```

# Extracting climate data

The 'raster' library has a function for referencing the climate data stored in our raster for every occurrence point. These data are the primary input to our species distribution models.

```{r}
extr = extract(wc, sp_df[,c('longitude', 'latitude')])
summary(extr)
extr=as.data.frame(na.omit(extr))

ggplot(data=extr) +
  geom_histogram(aes(x=bio12), bins=30) + 
  theme_bw()

```


### Challenge/Homework:

Choose a species observed in the [Stonehill Biodiversity iNaturalist Project](https://www.inaturalist.org/observations?project_id=stonehill-biodiversity) Query GBIF using 'spocc' for occurrence data records. Create a ggplot based map using bio1 as a background to show the distribution of this species.

Post the script and your map to your GitHub repo (you may want to create a repository for this course on your account). When that is up post a link to your repository on Slack using the [biodiversity](https://introbioinformatics.slack.com/archives/CQ5NRTR9R) channel. 

## Reading:

By Monday, read the [ENMeval](https://besjournals.onlinelibrary.wiley.com/doi/pdf/10.1111/2041-210X.12261) paper. To go with this write a ~1 page reflection on the paper and submit that by email before class.

[home](https://rsh249.github.io/bioinformatics)

