---
title: "Spatial Data in R"
author: "Prof. Harbert"
date: "November 12, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installing necessary packages

For the next few lessons you will need:

```{r, eval=FALSE}
install.packages('dismo') 
install.packages('raster')
install.packages('spocc')
install.packages('ENMeval')

#And maybe

devtools::install_github('rsh249/vegdistmod')


```

### Review data sources

We will be getting data for distribution modeling from:

+ [WorldClim](https://www.worldclim.org)
+ [GBIF](https://www.gbif.org)
+ [iNaturalist](https://www.inaturalist.org)

# Geographic data: Continuous surfaces -- > Raster data objects

Geographic data that are continuous across your study area can be represented as a gridded image-like data type known as a raster. These are basically really big 2 dimensional tables but with special attributes that help align the values in the matrix across the curved surface of Earth.

Our primary raster data will be climate. We will get that from the WorldClim project.

### Downloading climate data: WorldClim

We can download climate data for land masses except Antarctica using the 'raster' library and:

```{r}
library(raster)
clim = getData('worldclim', var='bio', res=5)

summary(clim)
summary(clim[[1]])
plot(clim[[1]], col= topo.colors(9))
```

See [here](https://www.worldclim.org/bioclim) for Bioclim variable definitions.

We are looking at 'bio1' above which is the mean annual temperature value in Celsius degrees * 10.

### Raster operations: Crop

If we want to 'zoom in' on a region of interest we can crop the raster just like a picture file. However, we must use the coordinates of the raster. In this case that is latitude and longitude in decimal degrees.

```{r}

#define an 'extent' object.
#eastern US
ext = extent(-100, -50, 20, 55)

#crop
c2 = crop(clim, ext)
plot(c2[[1]], col=topo.colors(99))

```


# Next time: Vectors data, mining GBIF/iNaturalist, and ENMeval for distribution modeling.

[Home](https://rsh249.github.io/bioinformatics)